Prediction of PIK3CA mutation with gene expression in cancer

### 2020-05-27 plosOne/plosOne.Rmd
first draft

---
title: Prediction of PIK3CA mutation with gene expression
author:
  - name: Jun Kang
    email: jkang.alien@gmail.com
    affiliation: Department of Hospital Pathology, Seoul St. Mary’s Hospital, College of Medicine, The Catholic University of Korea, Seoul, South Korea

  - name: Ahwon Lee
    email: klee@catholic.ac.kr
    affiliation: Department of Hospital Pathology, Seoul St. Mary’s Hospital, College of Medicine, The Catholic University of Korea, Seoul, South Korea

  - name: Youn Soo Lee
    email: lys9908@catholic.ac.kr
    affiliation: Department of Hospital Pathology, Seoul St. Mary’s Hospital, College of Medicine, The Catholic University of Korea, Seoul, South Korea
    corresponding: lys9908@catholic.ac.kr

abstract: | 
  Breast cancer with PIK3CA mutation has been approved to use PIK3CA inhibitor in hormone receptor-positive HER2 negative subtype. We apply a supervised elastic net penalized logistic regression model in prediction PIK3CA mutation from gene expression data. We used the TCGA pan-cancer dataset. Penalized logistic regression was applied to prediction modeling. 10-fold cross-validation with targe variable stratification was done over the hyperparameter grid. 10845 cases were available for both PIK3CA mutation and mRNA expression data. In 10-fold cross-validation, the model with $\lambda$ = 0.01 and $\alpha$ = 1.0 (Ridge regression) showed best performance in terms of AUROC. The final model was trained with the selected hyperparameters with the entire trainset. The trainset AUROC was 0.93 and the testset AUROC was 0.84. The AUPR of trainset was 0.66 and the testset AUPR was 0.39. Both IGFR1R and PTEN are key players in the tyrosine kinase pathway. The cancer type was an important predictor. Our study suggests that the prediction of genomic alteration with gene expression data is possible with good performance.

@@

Targeted therapy becomes a standard treatment in many cancer patients. Much targeted therapy requires a test for a specific cancer genomic alteration, to treat the patients. Many direct tests for the genomic alteration have been developed and prooved for their clinical utility to find which patients will be responding to the targeted therapy.

@@

Prediction RAS pathway activation with gene expression data was done in the previous study. @way_2018_machine_cellreports They trained pan-cancer The Cancer Genome Atlas (TCGA) data with a supervised elastic net penalized logistic regression classifier with stochastic gradient descent. The performance of their model was 84% with an area under the receiver operating characteristic (AUROC) curve and 63% with an area under the precision-recall (AUPR) curve. The authors suggested their approach can be applied to other genomic alterations.

@@

We apply a supervised elastic net penalized logistic regression model in prediction PIK3CA mutation. The purpose of this study is to know this prediction model approach can be applied not only to RAS pathway activation but also PIK3CA mutation across many cancer types.

@@

We used the TCGA pan-cancer dataset. TCGA is a cancer genomic consortium that archives data of exome sequencing, gene expression, DNA methylation, protein expression, and clinical data of more than 10000 cancer samples across 33 common cancer types. The gene expression TCGA pan-cancer dataset is batch-corrected with normalization. The TCGA dataset is publically available. PIK3CA mutation data was got using cgdsr rpackage.[@jacobsenCgdsrRbasedAPI2019] Gene expression data was downloaded from the National Cancer Institute (NCI)'s Genomic Data Commons (GDC) website that archives data used for The Pan-Cancer Atlas initiative. (https://gdc.cancer.gov/about-data/publications/pancanatlas)

The target variable was the PIK3CA mutation status. The status of  PIK3CA was considered as positive when the case has following PIK3CA variants which are the target variables of the therascreen PIK3CA RGQ PCR Kit; C420R, E542K, E545A, E545D, E545G, E545K, Q546E, Q546R, H1047L, H1047R, H1047Y. The therascreen PIK3CA RGQ PCR Kit was approved as a companion diagnostics to treat with PIK3CA inhibitor by the U.S. Food and Drug Administration. We split the three-quarters of the dataset for the trainset and one quarter for the test set.

@@

Model performance was evaluated with the area under the receiver operating characteristic (ROC) and area under the ROC (AUROC) and precision-recall (PR) and area under the curve (AUPR). The AUPR is more informative than AUROC on imbalanced datasets. [@saitoPrecisionRecallPlotMore2015] The modeling process and accessing model performance were done with tidymodels rpackage.  [@kuhnTidymodelsEasilyInstall2020]

@@

Figure 3 shows the top 30 important predictors. The coefficient is the parameter of the predictor which represents the effect of the predictor on prediction. IGF1R mRNA expression was the strongest negative predictor and PTEN was the strongest positive predictor. Both IGFR1R and PTEN are key players in the tyrosine kinase pathway. The cancer type was an important predictor. Some cancer types including uterine carcinosarcoma (UCS), bladder urothelial carcinoma (BLCA), pancreatic adenocarcinoma (PAAD), lymphoid neoplasm diffuse large B-cell lymphoma (DLBC) are the strongest predictors.

@@

Prediction modeling from the TCGA pan-cancer dataset has limitations regarding data preprocessing. Methods for cancer gene expression have been developed for research. The gene expression data is processed between-sample normalization to remove the batch effect. If the model has been trained from the between-sample normalization, a new sample can not be exactly preprocessed with normalization which was done on trainset. A model based on gene expression TCGA pan-cancer dataset has a limitation on data preprocessing. It is necessary for developing a preprocessing method that is independent with a dataset to apply the gene expression data to the prediction model.

@@

Our model includes cancer type predictor and they are stronger than gene expression data. The varying prevalence of PIC3CA mutation across cancer type might reason for the strong cancer type predictor. If the cancer type was wrong or can not be determined, our model performance can be poor.

Some significant gene expression predictors were closely related to PTEN and the PI3K pathway. PTEN and IGFR1R are the strongest gene expression predictor which have negative and positive predictive power. IGF1R is a tyrosine kinase receptor that activates PI3K. [@leroithInsulinlikeGrowthFactor2003] Insulin receptor substrate-2 (IRS2) is the adaptor protein of IGF1R. [@heInteractionInsulinReceptor1996] PTEN is an important regulator of PIP~3~ by dephosphorylating  PIP~3~ in contrast PI3K.[@cantleyPhosphoinositide3KinasePathway2002]

@@

- Figure 1. prevalence rate of PIK3CA across cancer types
The abbreviations of cancer types are explained in the S1 Appendix.

@@

(A) Left: receiver operating characteristic (ROC) curve right: precision-recall (PR) curve of trainset and test set. The horizontal green line is the PIK3CA mutation rate (0.11)  (B) Correlation between trainset and test set of the area under the receiver operating characteristic curve (AUROC) and area under the precision-recall curve (AUPR) among cancer type. The gray band is 95% confidence interval. The abbreviations are explained in the S1 Appendix. (C) Correlation between the PIK3CA mutation rate of area under the receiver operating characteristic curve (AUROC) and area under the precision-recall curve (AUPR).

>>>>

---
title: Prediction of PIK3CA mutation with gene expression in cancer
author:
  - name: Jun Kang
    email: jkang.alien@gmail.com
    affiliation: Department of Hospital Pathology, Seoul St. Mary’s Hospital
  - name: Ahwon Lee
    email: klee@catholic.ac.kr
    affiliation: Department of Hospital Pathology, Seoul St. Mary’s Hospital
  - name: Youn Soo Lee
    email: lys9908@catholic.ac.kr
    affiliation: Department of Hospital Pathology, Seoul St. Mary’s Hospital
    corresponding: lys9908@catholic.ac.kr
address:
  - code: Department of Hospital Pathology, Seoul St. Mary’s Hospital
    address: Department of Hospital Pathology, Seoul St. Mary’s Hospital, College of Medicine, The Catholic University of Korea, Seoul, South Korea

abstract: | 
  Breast cancer with PIK3CA mutation has been approved to use PIK3CA inhibitor in hormone receptor-positive HER2 negative subtype. We apply a supervised elastic net penalized logistic regression model in prediction PIK3CA mutation from gene expression data. Penalized logistic regression was applied to prediction modeling using the TCGA pan-cancer dataset. 10845 cases were available for both PIK3CA mutation and mRNA expression data. In 10-fold cross-validation, the model with $\lambda$ = 0.01 and $\alpha$ = 1.0 (Ridge regression) showed best performance in terms of AUROC. The final model was trained with the selected hyperparameters with the entire trainset. The trainset AUROC was 0.93 and the testset AUROC was 0.84. The AUPR of trainset was 0.66 and the testset AUPR was 0.39. The cancer types were the most important predictors. Both IGFR1R and PTEN were most important among gene expression predictors. Our study suggests that the prediction of genomic alteration with gene expression data is possible with good performance.

@@

Targeted therapy becomes a standard treatment in many cancer patients. The majority of targeted therapy requires a test for a specific cancer genomic alteration, to treat the patients. Many direct tests for the genomic alteration have been developed and prooved for their clinical utility to find which patients will be responding to the targeted therapy.

@@

Prediction of RAS pathway activation with gene expression data was done in the previous study. @way_2018_machine_cellreports They trained pan-cancer The Cancer Genome Atlas (TCGA) data with a supervised elastic net penalized logistic regression classifier with stochastic gradient descent. The performance of their model was 84% with an area under the receiver operating characteristic (AUROC) curve and 63% with an area under the precision-recall (AUPR) curve. The authors suggested their approach can be applied to other genomic alterations.

@@

We apply a supervised elastic net penalized logistic regression model in prediction PIK3CA mutation. The purpose of this study is to know whether this prediction model approach can be applied not only to RAS pathway activation but also PIK3CA mutation across many cancer types.

@@

We used the TCGA pan-cancer dataset. TCGA is a cancer genomic consortium that archives data of exome sequencing, gene expression, DNA methylation, protein expression, and clinical data of more than 10000 cancer samples across 33 common cancer types. The TCGA dataset is publically available. PIK3CA mutation data was got using cgdsr rpackage.[@jacobsenCgdsrRbasedAPI2019] Gene expression data was downloaded from the National Cancer Institute (NCI)'s Genomic Data Commons (GDC) website that archives data used for The Pan-Cancer Atlas initiative. (https://gdc.cancer.gov/about-data/publications/pancanatlas) The gene expression TCGA pan-cancer dataset is batch-corrected with normalization.

The target variable was the PIK3CA mutation status. The status of  PIK3CA was considered as positive when the case has following PIK3CA variants which are the target variants of the therascreen PIK3CA RGQ PCR Kit; C420R, E542K, E545A, E545D, E545G, E545K, Q546E, Q546R, H1047L, H1047R, H1047Y. The therascreen PIK3CA RGQ PCR Kit was approved as a companion diagnostics to treat with PIK3CA inhibitor by the U.S. Food and Drug Administration. We split the three-quarters of the dataset for the trainset and one quarter for the test set.

@@

Model performance was evaluated with the AUROC curve and the AUPR curve. The AUPR is more informative than AUROC on imbalanced datasets. [@saitoPrecisionRecallPlotMore2015] The modeling process and accessing model performance were done with tidymodels rpackage.  [@kuhnTidymodelsEasilyInstall2020]

@@

Figure 3 shows the top 30 important predictors. The coefficient is the parameter of the predictor which represents the effect of the predictor on prediction. IGF1R mRNA expression was the strongest negative predictor and PTEN was the strongest positive predictor. Both IGFR1R and PTEN are key players in the tyrosine kinase pathway. The cancer types were important predictors. Some cancer types including uterine carcinosarcoma (UCS), bladder urothelial carcinoma (BLCA), pancreatic adenocarcinoma (PAAD), lymphoid neoplasm diffuse large B-cell lymphoma (DLBC) are the strongest predictors.

@@

Prediction modeling from the TCGA pan-cancer dataset has limitations regarding data preprocessing. The gene expression data is processed between-sample normalization to remove the batch effect. If the model has been trained from the between-sample normalization, a new sample can not be exactly preprocessed with normalization which was done on trainset. A model based on gene expression TCGA pan-cancer dataset has a limitation on data preprocessing. It is necessary for developing a preprocessing method that is independent with a dataset to apply the gene expression data to the prediction model.

@@

Our model includes cancer type as predictors and they are stronger predictors than gene expression. The varying prevalence of PIC3CA mutation across cancer type might reason for the strong predictive power of cancer type. If the cancer type was wrong or can not be determined, our model performance can be poor.

Some significant gene expression predictors were closely related to PTEN and the PI3K pathway. PTEN and IGFR1R are the strongest gene expression predictor which have negative and positive predictive power. IGF1R is a tyrosine kinase receptor that activates PI3K. [@leroithInsulinlikeGrowthFactor2003] PTEN is an important regulator of PIP~3~ by dephosphorylating  PIP~3~ in contrast PI3K.[@cantleyPhosphoinositide3KinasePathway2002]

@@

- Figure 1. Prevalence rate of PIK3CA across cancer types.
The abbreviations of cancer types are explained in the S1 Appendix.

@@

(A) Left: receiver operating characteristic (ROC) curve. right: precision-recall (PR) curve of trainset and test set. The horizontal green line is the PIK3CA mutation rate (0.11)  (B) Correlation between trainset and test set of the area under the receiver operating characteristic curve (AUROC) and the area under the precision-recall curve (AUPR) among cancer type. The gray band is 95% confidence interval. The abbreviations are explained in the S1 Appendix. (C) Correlation between the PIK3CA mutation rate of the AUROC and the AUPR.

### 2020-05-27 plosOne/plosOne.tex
first draft

\newcommand{\getAff}[1]{
  \getIndex{#1}{}
}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Prediction of PIK3CA mutation with gene expression} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jun Kang\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital, College of
Medicine, The Catholic University of Korea, Seoul, South Korea}},
Ahwon Lee\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital, College of
Medicine, The Catholic University of Korea, Seoul, South Korea}},
Youn Soo Lee\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital, College of
Medicine, The Catholic University of Korea, Seoul, South Korea}}\textsuperscript{*}\\
\bigskip
\bigskip
* Corresponding author: lys9908@catholic.ac.kr\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Breast cancer with PIK3CA mutation has been approved to use PIK3CA
inhibitor in hormone receptor-positive HER2 negative subtype. We apply a
supervised elastic net penalized logistic regression model in prediction
PIK3CA mutation from gene expression data. We used the TCGA pan-cancer
dataset. Penalized logistic regression was applied to prediction
modeling. 10-fold cross-validation with targe variable stratification
was done over the hyperparameter grid. 10845 cases were available for
both PIK3CA mutation and mRNA expression data. In 10-fold
cross-validation, the model with \(\lambda\) = 0.01 and \(\alpha\) = 1.0
(Ridge regression) showed best performance in terms of AUROC. The final
model was trained with the selected hyperparameters with the entire
trainset. The trainset AUROC was 0.93 and the testset AUROC was 0.84.
The AUPR of trainset was 0.66 and the testset AUPR was 0.39. Both IGFR1R
and PTEN are key players in the tyrosine kinase pathway. The cancer type
was an important predictor. Our study suggests that the prediction of
genomic alteration with gene expression data is possible with good
performance.

@@

Prediction RAS pathway activation with gene expression data was done in
the previous study. {[}5{]} They trained pan-cancer The Cancer Genome
Atlas (TCGA) data with a supervised elastic net penalized logistic
regression classifier with stochastic gradient descent. The performance
of their model was 84\% with an area under the receiver operating
characteristic (AUROC) curve and 63\% with an area under the
precision-recall (AUPR) curve. The authors suggested their approach can
be applied to other genomic alterations.

PIK3CA encodes the p110\(\alpha\) catalytic subunit of
phosphatidylinositol 3′-kinase (PI3K). PI3K is a protein kinase that
phosphorylates phosphatidylinositol 4,5-biphosphate
(PIP\textsubscript{2}) to make phosphatidylinositol 3,4,5-triphosphate
(PIP\textsubscript{3}). Phosphatase and tensin homolog (PTEN) changes
PIP\textsubscript{2} to PIP\textsubscript{3} in contrast, PI3K.
PIP\textsubscript{3} is a second messenger to activate protein kinase B
(AKT) which is a serine/threonine-specific protein kinase. AKT inhibits
apoptosis and promotes cell proliferation. {[}6{]}

Breast cancer with PIK3CA mutation has been approved to use PIK3CA
inhibitor in hormone receptor-positive HER2 negative subtype. {[}7{]}
The PIK3CA mutation is the second most driver mutation after TP53. The
PIK3CA mutation is most frequently founded in endometrial carcinoma
(45\%), and followed by breast invasive carcinoma (24\%), cervical
squamous cell carcinoma, and endocervical adenocarcinoma (20\%) and
colon adenocarcinoma (16\%).

We apply a supervised elastic net penalized logistic regression model in
prediction PIK3CA mutation. The purpose of this study is to know this
prediction model approach can be applied not only to RAS pathway
activation but also PIK3CA mutation across many cancer types.

@@

We used the TCGA pan-cancer dataset. TCGA is a cancer genomic consortium
that archives data of exome sequencing, gene expression, DNA
methylation, protein expression, and clinical data of more than 10000
cancer samples across 33 common cancer types. The gene expression TCGA
pan-cancer dataset is batch-corrected with normalization. The TCGA
dataset is publically available. PIK3CA mutation data was got using
cgdsr rpackage.{[}8{]} Gene expression data was downloaded from the
National Cancer Institute (NCI)'s Genomic Data Commons (GDC) website
that archives data used for The Pan-Cancer Atlas initiative.
(https://gdc.cancer.gov/about-data/publications/pancanatlas)

The target variable was the PIK3CA mutation status. The status of PIK3CA
was considered as positive when the case has following PIK3CA variants
which are the target variables of the therascreen PIK3CA RGQ PCR Kit;
C420R, E542K, E545A, E545D, E545G, E545K, Q546E, Q546R, H1047L, H1047R,
H1047Y. The therascreen PIK3CA RGQ PCR Kit was approved as a companion
diagnostics to treat with PIK3CA inhibitor by the U.S. Food and Drug
Administration. We split the three-quarters of the dataset for the
trainset and one quarter for the test set.

@@

Model performance was evaluated with the area under the receiver
operating characteristic (ROC) and area under the ROC (AUROC) and
precision-recall (PR) and area under the curve (AUPR). The AUPR is more
informative than AUROC on imbalanced datasets. {[}11{]} The modeling
process and accessing model performance were done with tidymodels
rpackage. {[}12{]}

@@

Our model includes cancer type predictor and they are stronger than gene
expression data. The varying prevalence of PIC3CA mutation across cancer
type might reason for the strong cancer type predictor. If the cancer
type was wrong or can not be determined, our model performance can be
poor.

Some significant gene expression predictors were closely related to PTEN
and the PI3K pathway. PTEN and IGFR1R are the strongest gene expression
predictor which have negative and positive predictive power. IGF1R is a
tyrosine kinase receptor that activates PI3K. {[}13{]} Insulin receptor
substrate-2 (IRS2) is the adaptor protein of IGF1R. {[}14{]} PTEN is an
important regulator of PIP\textsubscript{3} by dephosphorylating
PIP\textsubscript{3} in contrast PI3K.{[}6{]}

There is some study trying to predict genomic alterations from gene
expression data. A study performed PIK3CA mutation prediction by a
gene-expression signature which is a sum of the average of the
logarithmic gene expression. The model showed good performance AUROC
0.71 in an independent test set. {[}4,15{]} Another study predicted copy
number alterations with gene expression using a multinomial logistic
regression model with least absolute shrinkage and selection operator
(LASSO). The prediction of the 1p/19q codel was very good with an AUROC
of 0.997. The gene-level prediction was good with an AUROC 0.75.
{[}16{]} A Hidden Markov Model Approach for Prediction of Genomic
Alterations from Gene Expression Profiling.{[}17{]} A logistic
regression model was used for MYCN gene amplification in neuroblastoma.
{[}18{]}

@@

\begin{itemize}
\item
  Figure 1. prevalence rate of PIK3CA across cancer types The
  abbreviations of cancer types are explained in the S1 Appendix.
\item
  Figure 2. Summary of modeling results
\end{itemize}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\tightlist
\item
  Left: receiver operating characteristic (ROC) curve right:
  precision-recall (PR) curve of trainset and test set. The horizontal
  green line is the PIK3CA mutation rate (0.11) (B) Correlation between
  trainset and test set of the area under the receiver operating
  characteristic curve (AUROC) and area under the precision-recall curve
  (AUPR) among cancer type. The gray band is 95\% confidence interval.
  The abbreviations are explained in the S1 Appendix. (C) Correlation
  between the PIK3CA mutation rate of area under the receiver operating
  characteristic curve (AUROC) and area under the precision-recall curve
  (AUPR).
\end{enumerate}

@@

\leavevmode\hypertarget{ref-cantleyPhosphoinositide3KinasePathway2002}{}%
6. Cantley LC. The Phosphoinositide 3-Kinase Pathway. Science. American
Association for the Advancement of Science; 2002;296: 1655--1657.
doi:\href{https://doi.org/10.1126/science.296.5573.1655}{10.1126/science.296.5573.1655}

\leavevmode\hypertarget{ref-andre_2019_alpelisib_nengljmeda}{}%
7. André F, Ciruelos E, Rubovszky G, Campone M, Loibl S, Rugo HS, et al.
Alpelisib for PIK3CA-Mutated, Hormone ReceptorPositive Advanced Breast
Cancer. New England Journal of Medicine. Massachusetts Medical Society;
2019;380: 1929--1940.
doi:\href{https://doi.org/10.1056/NEJMoa1813904}{10.1056/NEJMoa1813904}

@@

\leavevmode\hypertarget{ref-heInteractionInsulinReceptor1996}{}%
14. He W, Craparo A, Zhu Y, O'Neill TJ, Wang L-M, Pierce JH, et al.
Interaction of Insulin Receptor Substrate-2 (IRS-2) with the Insulin and
Insulin-like Growth Factor I Receptors EVIDENCE FOR TWO DISTINCT
PHOSPHOTYROSINE-DEPENDENT INTERACTION DOMAINS WITHIN IRS-2. J Biol Chem.
American Society for Biochemistry and Molecular Biology; 1996;271:
11641--11645.
doi:\href{https://doi.org/10.1074/jbc.271.20.11641}{10.1074/jbc.271.20.11641}

\leavevmode\hypertarget{ref-cizkovaGeneExpressionProfiling2010}{}%
15. Cizkova M, Cizeron-Clairac G, Vacher S, Susini A, Andrieu C,
Lidereau R, et al. Gene Expression Profiling Reveals New Aspects of
PIK3CA Mutation in ERalpha-Positive Breast Cancer: Major Implication of
the Wnt Signaling Pathway. PLOS ONE. Public Library of Science; 5:
e15647.
doi:\href{https://doi.org/10.1371/journal.pone.0015647}{10.1371/journal.pone.0015647}

\leavevmode\hypertarget{ref-muCNAPEMachineLearning2019}{}%
16. Mu Q, Wang J. CNAPE: A Machine Learning Method for Copy Number
Alteration Prediction from Gene Expression. IEEE/ACM Transactions on
Computational Biology and Bioinformatics. 2019; 1--1.
doi:\href{https://doi.org/10.1109/TCBB.2019.2944827}{10.1109/TCBB.2019.2944827}

\leavevmode\hypertarget{ref-gengHiddenMarkovModel2008}{}%
17. Geng H, Ali HH, Chan WC. A Hidden Markov Model Approach for
Prediction of Genomic Alterations from Gene Expression Profiling. In:
Măndoiu I, Sunderraman R, Zelikovsky A, editors. Bioinformatics Research
and Applications. Berlin, Heidelberg: Springer; 2008. pp. 414--425.
doi:\href{https://doi.org/10.1007/978-3-540-79450-9_38}{10.1007/978-3-540-79450-9\_38}

\leavevmode\hypertarget{ref-heGeneSignaturesAssociated2020}{}%
18. He X, Qin C, Zhao Y, Zou L, Zhao H, Cheng C. Gene signatures
associated with genomic aberrations predict prognosis in neuroblastoma.
Cancer Communications. 2020;40: 105--118.
doi:\href{https://doi.org/10.1002/cac2.12016}{10.1002/cac2.12016}

>>>>

\newcommand{\getAff}[1]{
  \getIndex{#1}{Department of Hospital Pathology, Seoul St.~Mary's Hospital}
}

@@

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Prediction of PIK3CA mutation with gene expression in cancer} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Jun Kang\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}},
Ahwon Lee\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}},
Youn Soo Lee\textsuperscript{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}}\textsuperscript{*}\\
\bigskip
\textbf{\getAff{Department of Hospital Pathology, Seoul St.~Mary's Hospital}}Department of Hospital Pathology, Seoul St.~Mary's Hospital, College of
Medicine, The Catholic University of Korea, Seoul, South Korea\\
\bigskip
* Corresponding author: lys9908@catholic.ac.kr\\
\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Breast cancer with PIK3CA mutation has been approved to use PIK3CA
inhibitor in hormone receptor-positive HER2 negative subtype. We apply a
supervised elastic net penalized logistic regression model in prediction
PIK3CA mutation from gene expression data. Penalized logistic regression
was applied to prediction modeling using the TCGA pan-cancer dataset.
10845 cases were available for both PIK3CA mutation and mRNA expression
data. In 10-fold cross-validation, the model with \(\lambda\) = 0.01 and
\(\alpha\) = 1.0 (Ridge regression) showed best performance in terms of
AUROC. The final model was trained with the selected hyperparameters
with the entire trainset. The trainset AUROC was 0.93 and the testset
AUROC was 0.84. The AUPR of trainset was 0.66 and the testset AUPR was
0.39. The cancer types were the most important predictors. Both IGFR1R
and PTEN were most important among gene expression predictors. Our study
suggests that the prediction of genomic alteration with gene expression
data is possible with good performance.

@@

Prediction of RAS pathway activation with gene expression data was done
in the previous study. {[}5{]} They trained pan-cancer The Cancer Genome
Atlas (TCGA) data with a supervised elastic net penalized logistic
regression classifier with stochastic gradient descent. The performance
of their model was 84\% with an area under the receiver operating
characteristic (AUROC) curve and 63\% with an area under the
precision-recall (AUPR) curve. The authors suggested their approach can
be applied to other genomic alterations.

Breast cancer with PIK3CA mutation has been approved to use PIK3CA
inhibitor in hormone receptor-positive HER2 negative subtype. {[}6{]}
The PIK3CA mutation is the second most driver mutation after TP53. The
PIK3CA mutation is most frequently founded in endometrial carcinoma
(45\%), and followed by breast invasive carcinoma (24\%), cervical
squamous cell carcinoma, and endocervical adenocarcinoma (20\%) and
colon adenocarcinoma (16\%).

PIK3CA encodes the p110\(\alpha\) catalytic subunit of
phosphatidylinositol 3′-kinase (PI3K). PI3K is a protein kinase that
phosphorylates phosphatidylinositol 4,5-biphosphate
(PIP\textsubscript{2}) to make phosphatidylinositol 3,4,5-triphosphate
(PIP\textsubscript{3}). Phosphatase and tensin homolog (PTEN) changes
PIP\textsubscript{2} to PIP\textsubscript{3} in contrast, PI3K.
PIP\textsubscript{3} is a second messenger to activate protein kinase B
(AKT) which is a serine/threonine-specific protein kinase. AKT inhibits
apoptosis and promotes cell proliferation. {[}7{]}

We apply a supervised elastic net penalized logistic regression model in
prediction PIK3CA mutation. The purpose of this study is to know whether
this prediction model approach can be applied not only to RAS pathway
activation but also PIK3CA mutation across many cancer types.

@@

We used the TCGA pan-cancer dataset. TCGA is a cancer genomic consortium
that archives data of exome sequencing, gene expression, DNA
methylation, protein expression, and clinical data of more than 10000
cancer samples across 33 common cancer types. The TCGA dataset is
publically available. PIK3CA mutation data was got using cgdsr
rpackage.{[}8{]} Gene expression data was downloaded from the National
Cancer Institute (NCI)'s Genomic Data Commons (GDC) website that
archives data used for The Pan-Cancer Atlas initiative.
(https://gdc.cancer.gov/about-data/publications/pancanatlas) The gene
expression TCGA pan-cancer dataset is batch-corrected with
normalization.

The target variable was the PIK3CA mutation status. The status of PIK3CA
was considered as positive when the case has following PIK3CA variants
which are the target variants of the therascreen PIK3CA RGQ PCR Kit;
C420R, E542K, E545A, E545D, E545G, E545K, Q546E, Q546R, H1047L, H1047R,
H1047Y. The therascreen PIK3CA RGQ PCR Kit was approved as a companion
diagnostics to treat with PIK3CA inhibitor by the U.S. Food and Drug
Administration. We split the three-quarters of the dataset for the
trainset and one quarter for the test set.

@@

Model performance was evaluated with the AUROC curve and the AUPR curve.
The AUPR is more informative than AUROC on imbalanced datasets. {[}11{]}
The modeling process and accessing model performance were done with
tidymodels rpackage. {[}12{]}

@@

Our model includes cancer type as predictors and they are stronger
predictors than gene expression. The varying prevalence of PIC3CA
mutation across cancer type might reason for the strong predictive power
of cancer type. If the cancer type was wrong or can not be determined,
our model performance can be poor.

Some significant gene expression predictors were closely related to PTEN
and the PI3K pathway. PTEN and IGFR1R are the strongest gene expression
predictor which have negative and positive predictive power. IGF1R is a
tyrosine kinase receptor that activates PI3K. {[}13{]} PTEN is an
important regulator of PIP\textsubscript{3} by dephosphorylating
PIP\textsubscript{3} in contrast PI3K.{[}7{]}

There is some study trying to predict genomic alterations from gene
expression data. A study performed PIK3CA mutation prediction by a
gene-expression signature which is a sum of the average of the
logarithmic gene expression. The model showed good performance AUROC
0.71 in an independent test set. {[}4,14{]} Another study predicted copy
number alterations with gene expression using a multinomial logistic
regression model with least absolute shrinkage and selection operator
(LASSO). The prediction of the 1p/19q codel was very good with an AUROC
of 0.997. The gene-level prediction was good with an AUROC 0.75.
{[}15{]} A Hidden Markov Model Approach for Prediction of Genomic
Alterations from Gene Expression Profiling.{[}16{]} A logistic
regression model was used for MYCN gene amplification in neuroblastoma.
{[}17{]}

@@

\begin{itemize}
\item
  Figure 1. Prevalence rate of PIK3CA across cancer types. The
  abbreviations of cancer types are explained in the S1 Appendix.
\item
  Figure 2. Summary of modeling results
\end{itemize}

\begin{enumerate}
\def\labelenumi{(\Alph{enumi})}
\tightlist
\item
  Left: receiver operating characteristic (ROC) curve. right:
  precision-recall (PR) curve of trainset and test set. The horizontal
  green line is the PIK3CA mutation rate (0.11) (B) Correlation between
  trainset and test set of the area under the receiver operating
  characteristic curve (AUROC) and the area under the precision-recall
  curve (AUPR) among cancer type. The gray band is 95\% confidence
  interval. The abbreviations are explained in the S1 Appendix. (C)
  Correlation between the PIK3CA mutation rate of the AUROC and the
  AUPR.
\end{enumerate}

@@

\leavevmode\hypertarget{ref-andre_2019_alpelisib_nengljmeda}{}%
6. André F, Ciruelos E, Rubovszky G, Campone M, Loibl S, Rugo HS, et al.
Alpelisib for PIK3CA-Mutated, Hormone ReceptorPositive Advanced Breast
Cancer. New England Journal of Medicine. Massachusetts Medical Society;
2019;380: 1929--1940.
doi:\href{https://doi.org/10.1056/NEJMoa1813904}{10.1056/NEJMoa1813904}

\leavevmode\hypertarget{ref-cantleyPhosphoinositide3KinasePathway2002}{}%
7. Cantley LC. The Phosphoinositide 3-Kinase Pathway. Science. American
Association for the Advancement of Science; 2002;296: 1655--1657.
doi:\href{https://doi.org/10.1126/science.296.5573.1655}{10.1126/science.296.5573.1655}

@@

\leavevmode\hypertarget{ref-cizkovaGeneExpressionProfiling2010}{}%
14. Cizkova M, Cizeron-Clairac G, Vacher S, Susini A, Andrieu C,
Lidereau R, et al. Gene Expression Profiling Reveals New Aspects of
PIK3CA Mutation in ERalpha-Positive Breast Cancer: Major Implication of
the Wnt Signaling Pathway. PLOS ONE. Public Library of Science; 5:
e15647.
doi:\href{https://doi.org/10.1371/journal.pone.0015647}{10.1371/journal.pone.0015647}

\leavevmode\hypertarget{ref-muCNAPEMachineLearning2019}{}%
15. Mu Q, Wang J. CNAPE: A Machine Learning Method for Copy Number
Alteration Prediction from Gene Expression. IEEE/ACM Transactions on
Computational Biology and Bioinformatics. 2019; 1--1.
doi:\href{https://doi.org/10.1109/TCBB.2019.2944827}{10.1109/TCBB.2019.2944827}

\leavevmode\hypertarget{ref-gengHiddenMarkovModel2008}{}%
16. Geng H, Ali HH, Chan WC. A Hidden Markov Model Approach for
Prediction of Genomic Alterations from Gene Expression Profiling. In:
Măndoiu I, Sunderraman R, Zelikovsky A, editors. Bioinformatics Research
and Applications. Berlin, Heidelberg: Springer; 2008. pp. 414--425.
doi:\href{https://doi.org/10.1007/978-3-540-79450-9_38}{10.1007/978-3-540-79450-9\_38}

\leavevmode\hypertarget{ref-heGeneSignaturesAssociated2020}{}%
17. He X, Qin C, Zhao Y, Zou L, Zhao H, Cheng C. Gene signatures
associated with genomic aberrations predict prognosis in neuroblastoma.
Cancer Communications. 2020;40: 105--118.
doi:\href{https://doi.org/10.1002/cac2.12016}{10.1002/cac2.12016}
